Choosing Color Palettes for Statistical Graphics

### 2006-10-16 trunk/vcd/paper/Color/draft.Rnw
finished draft version for colors paper
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/vcd@590 edb9625f-4e0d-4859-8d74-9fd3b1da38cb

\title{Choosing Color Palettes for Statistical Graphics}
\author{Achim Zeileis and Kurt Hornik\\Wirtschaftsuniversit\"at Wien}
\Plainauthor{Achim Zeileis, Kurt Hornik}
\Keywords{qualitative palette, sequential palette, diverging palette, HCL colors, HSV colors, perceptually-based color space}

@@

generate_png <- TRUE
@

First, we motivate why HCL colors \citep{vcd:Ihaka:2003} are suitable for choosing color palettes
by contrasting them with the more commonly implemented HSV colors.

Then, we describe strategies how color palettes for categorical and numerical
data can be chosen in this space. Following \cite{vcd:Brewer:1999}, we distinguish
three types of palettes: qualitative, sequential and diverging.
All palettes described are 
available in \proglang{R} \citep{vcd:R:2006} in the package \pkg{vcd} \citep{vcd:Meyer+Zeileis+Hornik:2006}
using the HCL color implementation from \pkg{colorspace} \citep{vcd:Ihaka:2004}.
Technical documentation to the \proglang{R} implementations along with a large
collection of examples is available via \verb/help("rainbow_hcl")/ that provides 
more comparisons between existing \proglang{R} palettes (based on HSV colors) and
the HCL color palettes.

%%FIXME: what to say about ColorBrewer

@@

For choosing color palettes, it is imperative to have an understanding
how colors are perceived. For this it is helpful to have an idea how
human color vision evolved. It has been hypothesized that it developed in three
distinct stages: 1.~perception of \textit{light/dark} contrasts (monochrome only),
2.~\textit{yellow/blue} contrasts (usually associated with our notion of warm/cold colors),
3.~\textit{green/red} contrasts (helpful for assessing the ripeness of fruit). See 
\cite{vcd:Ihaka:2003} for more details and references.

Due to these three color axes, colors are typically described as
locations in 3-dimensional spaces. However, human perception of color
does not correspond to the physiological axes above, but rather to polar
coordinates in the color plane (yellow/blue vs.~green/red) plus a third
light/dark axis. Thus, perceptually-based color spaces are defined by three dimensions
that try to capture
\begin{enumerate}
  \item \textbf{hue} (dominant wavelength)
  \item \textbf{chroma} (colorfulness)
  \item \textbf{luminance} (brightness, amount of light)
\end{enumerate}

@@

A popular implementation of such a color space, available in many graphics and
statistics software packages, are \textit{HSV} (hue, saturation, value) colors.
They are a simple transformation of \textit{RGB} (red, green, blue) colors
and are defined by a triplet $(H, S, V)$ with $H \in [0, 360]$ and $S, V \in [0, 100]$.
HSV space has the shape of a single regular cone (often inflated to a regular
cylinder). Vertical sections through this space are shown in the upper panel
of Figure~\ref{fig:hsv-hcl}, depicting hue and saturation given value.
Although simple to specify and easily available in many computing environments,
HSV colors have a fundamental drawback: its three dimensions map to the three
dimensions of human color perception very poorly. The three dimensions are
confounded which is most easily seen when converting the vertical sections to 
gray scale images in Figure~\ref{fig:hsv-hcl-gray}. Clearly, the brightness of
colors is not uniform over hues and saturations (given value)---therefore,
HSV colors are often not considered to be perceptually based.

To overcome these drawbacks, various color spaces have been suggested
that properly map to the perception dimensions, the most prominent of which are
the CIELUV and CIELAB spaces developed by the \cite{vcd:CIE:2004}. \cite{vcd:Ihaka:2003}
argues that CIELUV colors are typically preferred for use with emissive technologies
such as computer screens which makes them an obvious candidate for implementation
in statistical software packages. By taking polar coordinates in the UV plane of
CIELUV, \textit{HCL} (hue, chroma, luminance) colors are obtained, defined by
a triplet $(H, C, L)$ with $H \in [0, 360]$ and $C, L \in [0, 100]$. HCL space
has the shape of a distorted double cone: the admissable chroma and luminance values
depend on the hue chosen. The lower panel of Figure~\ref{fig:hsv-hcl} shows
vertical sections through this space: each of the resulting hue/chroma planes (given
luminance) is now properly balanced towards the same gray (going from black to white
with increasing luminance) which becomes obvious when converting the colors to
a gray scale is in Figure~\ref{fig:hsv-hcl-gray}. This balancing of HCL colors
gives us the opportunity to conveniently choose color palettes which code categorical
and/or numerical information by translating it to the three perceptual dimensions.
However, some care is required for dealing with the irregular shape of the HCL space
which will be addressed in the following sections.

@@

\cite{vcd:Ihaka:2003} describes a simple strategy for choosing such palettes: chroma
and luminance are kept fixed and only the hue is varied for obtaining different colors
which are consequently all balanced towards the same gray. If colors from the full
color wheel (i.e., $H \in [0, 360]$) should be used, not all combinations of chroma
and luminance are feasible. Figure~\ref{fig:pal-q-construct} depicts how three colors
are chosen, given $C = 50$ and $L = 70$.

Various strategies for choosing the hues in a certain palette are conceivable.
A simple and intuitive one is to use colors as metaphors for categories
(e.g., for political parties). Figure~\ref{fig:pal-q-examples} shows a few
further examples for generating qualitative sets of colors $(H, 50, 70)$.
In the upper left panel colors from the full spectrum are used ($H = 30, 120, 210, 300$)
creating a `dynamic' set of colors. The upper right panel shows a `harmonic'
set with $H = 60, 120, 180, 240$. Warm colors (from the blue/green part of the spectrum: 
$H = 270, 230, 190, 150$) and cold colors (from the yellow/red part of the spectrum:
$H = 90, 50, 10, 330$) are shown in the lower left and right panel, respectively.

In \pkg{vcd}, these palettes are available in the function
\begin{Soutput}
rainbow_hcl(n, c = 50, l = 70, start = 0, end = 360*(n-1)/n, ...)
\end{Soutput}

%%FIXME: briefly explain interface

@@

@@

The simplest solution to this task is to employ light/dark contrasts, i.e., employ the
oldest and simplest perceptual axis. The interestingness is thus coded
by an increasing amount of gray (i.e., decreasing luminance) 
  \[ (H, 0, 90 - i \cdot 60), \]
where the hue $H$ used does not matter, chroma is set to 0 (i.e., no color), and 
luminance ranges in $[30, 90]$ avoiding the extreme colors white ($L = 100$) and 
black ($L = 0$). Instead of going linearly from light to dark gray, luminance
could also be increased nonlinearly, e.g., by some function $f(i)$ that
controls whether intensity/luminance is increased quickly or not. We found $f(i) = i^p$ 
to be a convenient transformation where the power $p$ can be varied to achieve
different degrees of non-linearity.

@@

\setkeys{Gin}{width=\textwidth}
\begin{figure}[b!]
\begin{center}
<<pal-s-construct>>=
if(generate_png) {
  png(file = "pal-s-construct.png", height = 420, width = 840)
  par(mfrow = c(1, 2))
  hue.slice(260, grid.n = 301)
  lines(c(0, 100), c(90, 50), lty = 2)
  lines(c(0, 80), c(90, 30), lty = 2)
  lines(c(2.5, 2.5), c(90, 30), lty = 2)
  rect(0, 87.5, 5, 92.5, col = hcl(260, 0, 90))
  rect(47.5, 67.5, 52.5, 72.5, col = hcl(260, 50, 70))
  rect(95, 47.5, 100, 52.5, col = hcl(260, 100, 50))
  rect(37.5, 57.5, 42.5, 62.5, col = hcl(260, 40, 60))
  rect(77.5, 27.5, 82.5, 32.5, col = hcl(260, 80, 30))
  rect(0, 57.5, 5, 62.5, col = hcl(260, 0, 60))
  rect(0, 27.5, 5, 32.5, col = hcl(260, 0, 30))
  hue.slice(120, grid.n = 301)
  lines(c(0, 100), c(90, 75), lty = 2)
  lines(c(2.5, 2.5), c(90, 30), lty = 2)
  rect(0, 87.5, 5, 92.5, col = hcl(260, 0, 90))
  rect(47.5, 80, 52.5, 85, col = hcl(120, 50, 82.5))
  rect(95, 72.5, 100, 77.5, col = hcl(120, 100, 75))
  rect(0, 57.5, 5, 62.5, col = hcl(260, 0, 60))
  rect(0, 27.5, 5, 32.5, col = hcl(260, 0, 30))
  dev.off()
}
@
\includegraphics{pal-s-construct}
\caption{\label{fig:pal-s-construct} Constructing sequential palettes. The chroma/luminance
plane is shown for two hues $H = 260$ (left) and $H = 120$ (right). Colors are chosen by varying
either only luminance or both luminance and chroma.}
\end{center}
\end{figure}

In \pkg{vcd}, this strategy is implemented in the function
\begin{Soutput}
sequential_hcl(n, h = 260, c = c(80, 0), l = c(30, 90), power = 1.5, ...) 
\end{Soutput}

\setkeys{Gin}{width=.8\textwidth}
\begin{figure}[p]
\begin{center}
<<pal-s-examples, fig=TRUE, height=9, width=6>>=
par(mfrow = c(5, 1), mar = c(0, 0, 0, 0))
pal(sequential_hcl(12, c = 0, power = 2.2))
pal(sequential_hcl(12, power = 2.2))
pal(terrain_hcl(12, c = c(65, 0), l = c(45, 95), power = c(1/3, 1.5)))
pal(heat_hcl(12, c = c(80, 30), l = c(30, 90), power = c(1/5, 1.5)))
pal(rev(heat_hcl(12, h = c(0, -100), l = c(75, 40), c = c(40, 80), power = 1)))
@
\caption{\label{fig:pal-s-examples} Examples for sequential palettes.}
\end{center}
\end{figure}

To increase the contrast between the colors in the palette even further, the ideas from
the previous sequential palettes can also be combined with qualitative palettes by
simultaneously varying the hue as well:
  \[ (H_2 - i \cdot (H_1 - H_2), C_{\max} - i^{p_1} \cdot (C_{\max} - C_{\min}), L_{\max} - i^{p_2} \cdot (L_{\max} - L_{\min})). \]
A typical application would be heat colors that increase from a light yellow (e.g., $(90, 30, 90)$)
to a full red (e.g., $(0, 100, 50)$). To make the change in hue visible, the chroma needs to
increase rather quickly for low values of $i$ and then only slowly for higher values of $i$.
This can be achieved by choosing a power $p_1 < 1$.

In \proglang{R}, these are available in the function
\begin{Soutput}
heat_hcl(n, h = c(0, 90), c = c(100, 30), l = c(50, 90), power = c(1/5, 1), ...)
\end{Soutput}
with which the lower three rows in Figure~\ref{fig:pal-s-examples} are produced.

@@

\setkeys{Gin}{width=\textwidth}
\begin{figure}[b!]
\begin{center}
<<pal-d-construct>>=
if(generate_png) {
  png(file = "pal-d-construct.png", height = 480, width = 540)
  hue260 <- hue.slice(260, grid.n = 301, plot = FALSE)
  hue360 <- hue.slice(360, grid.n = 301, plot = FALSE)
  mychroma <- as.numeric(colnames(hue260))
  luminance <- as.numeric(rownames(hue260))
  nc <- ncol(hue260)
  nl <- nrow(hue260)
  plot(0.5, 0.5, xlim = c(-100, 100), ylim = c(0, 100), type = "n", axes = FALSE,
       xlab = "", ylab = "", yaxs = "i", xaxs = "i", main = "")
  for(i in 1:(nc-1)) {
    rect(mychroma[i], luminance[-nl], mychroma[i] + 100/(nc-1), luminance[-1], border = hue260[,i+1], col = hue260[,i+1])
  }
  for(i in 1:(nc-1)) {
    rect(-mychroma[i], luminance[-nl], -(mychroma[i] + 100/(nc-1)), luminance[-1], border = hue360[,i+1], col = hue360[,i+1])
  }
  axis(1, labels = FALSE, at = c(-100, 0, 100))
  axis(2, at = c(0, 30, 50, 60, 70, 90, 100), labels = c(0, 30, 50, 60, 70, 90, 100))
  axis(3, at = -4:4*25, labels=c(4:0*25, 1:4*25))
  axis(4, at = c(0, 30, 50, 60, 70, 90, 100), labels = c(0, 30, 50, 60, 70, 90, 100))
  mtext(c("hue = 0", "hue = 260"), side = 3, at = c(-50, 50), line = 3, cex = 1.2)
  mtext("chroma", side = 3, at = 0, line = 2)
  mtext("luminance", side = 2, at = 50, line = 2)
  mtext("luminance", side = 4, at = 50, line = 2)
  lines(c(-100, 0, 100), c(50, 90, 50), lty = 2)
  rect(-100, 47.5, -90, 52.5, col = hcl(0, 100, 50))
  rect(-55, 67.5, -45, 72.5, col = hcl(0, 50, 70))
  rect(-5, 87.5, 5, 92.5, col = hcl(260, 0, 90))
  rect(45, 67.5, 55, 72.5, col = hcl(260, 50, 70))
  rect(90, 47.5, 100, 52.5, col = hcl(260, 100, 50))
  lines(c(-80, 0, 80), c(30, 90, 30), lty = 2)
  rect(-85, 27.5, -75, 32.5, col = hcl(0, 80, 30))
  rect(-45, 57.5, -35, 62.5, col = hcl(0, 40, 60))
  rect(-5, 87.5, 5, 92.5, col = hcl(260, 0, 90))
  rect(35, 57.5, 45, 62.5, col = hcl(260, 40, 60))
  rect(75, 27.5, 85, 32.5, col = hcl(260, 80, 30))
  dev.off()
}
@
\includegraphics{pal-d-construct}
\caption{\label{fig:pal-d-construct} Constructing diverging palettes. The chroma/luminance
plane is shown back to back hues $H = 0$ and $H = 260$. Colors are chosen by simultaneously 
varying luminance and chroma.}
\end{center}
\end{figure}

\setkeys{Gin}{width=.8\textwidth}
\begin{figure}[p]
\begin{center}
<<pal-d-examples, fig=TRUE, height=8, width=6>>=
par(mfrow = c(4, 1), mar = c(0, 0, 0, 0))
pal(diverge_hcl(7))
pal(diverge_hcl(7, c = 100, l = c(50, 90), power = 1))
pal(diverge_hcl(7, h = c(180, 330), c = 59, l = c(75, 95)))
pal(diverge_hcl(7, h = c(130, 43), c = 100, l = c(70, 90)))
@
\caption{\label{fig:pal-d-examples} Examples for diverging palettes.}
\end{center}
\end{figure}

Given sequential palettes, deriving diverging palettes is easy: two different hues are
chosen for adding color to the same amount of `gray' at a given intensity $|i|$.
Figure~\ref{fig:pal-d-construct} shows the chroma/luminance plane back to back for the
hues $H = 0$ and 260 with two different paths---with slightly different emphasis on
luminance or chroma contrasts---from a full red over a neutral grey to a full blue.
These particular hues were chosen because they have rather similar
chroma/luminance planes, allowing for many different combinations of maximal chroma and luminance.
%%FIXME: symmetric difference
As Figure~\ref{fig:pal-s-construct} illustrates, a bit of care is needed to choose two colors
that are sufficiently similar in their chroma/luminance plane to allow for the same
luminance and/or chroma contrasts.

In \pkg{vcd}, the following function is provided:
\begin{Soutput}
diverge_hcl(n, h = c(260, 0), c = 80, l = c(30, 90), power = 1.5, ...) 
\end{Soutput}

Figure~\ref{fig:pal-d-examples} shows various examples of conceivable combinations of
hue, chroma and luminance. The first palette uses a broader range on the luminance axis
whereas the others use larger ranges on the chroma axis.

\section{Illustrations}

>>>>

\title{Choosing Color Palettes for Statistical Graphics}
\author{Achim Zeileis \textnormal{and} Kurt Hornik\\Wirtschaftsuniversit\"at Wien}
\Plainauthor{Achim Zeileis, Kurt Hornik}
\Keywords{qualitative palette, sequential palette, diverging palette, HCL colors, HSV colors, perceptually-based color space}

@@

generate_png <- FALSE
@

Color is an integral element of graphical displays in general, and many statistical
graphics in particular. Statistical software packages typically provide various color
palettes and allow practitioners to employ these for visualizing data in various types
of displays. However, more often than not, there is relatively little guidance about
how to choose an appropriate palette for a certain visualization task. Here, we address
this problem by suggesting a color selection strategy for visualizing both categorical
and numerical information by selecting colors along axes in a color space whose axes
can be matched with perceptual axes of the human visual system.

For implementing color palettes in computer programs, typically color spaces are
employed that provide a mapping to Red-Green-Blue (RGB) colors. RGB is an additive
color model which is used for generating colors on computer screens
\citep[by mixing different amounts of the primary colors red, green and blue,
see][for more details]{vcd:Poynton:2000}. While being eminently useful for generating
colors on computer screens, it is hard for humans to select the coordinates in RGB
space corresponding to a certain color. Therefore, color picker tools in computer
programs are based on different color models---starting from the pioneering introduction
of Hue-Saturation-Value (HSV) colors \citep{vcd:Smith:1978}---which try to capture the
dimensions of human visual perception. HSV space is a simple transformation of RGB
space which is also implemented in many statistical software packages
\citep[see][for more details on HSV colors and links to further information about the other color
spaces discussed in the following]{vcd:Wiki+HSV:2006}. Although being easy to use,
HSV space exhibits a number of disadvantages: It is relatively difficult to select sets of
HSV coordinates that yield colors that are ``in harmony'' \citep[see][]{vcd:Munsell:1905}.
The reason for this is that colors with different hues also differ in saturation and brightness.
For statistical graphics, this is important because it can introduce size distortions in
the perception of shaded areas and color-caused optical illusions \citep{vcd:Cleveland+McGill:1983}.
Furthermore, the use of HSV colors encourages the use of flashy and highly saturated
colors which are good for drawing attention to a plot but hard to look at for a longer time.

Despite these known drawbacks, HSV space is still the predominantly-used color model in
color picker tools \citep{vcd:Moretti+Lyons:2002,vcd:Meier+Spalter+Karelitz:2004}
and the basis for many palettes in statistical software packages. A notable
exception is the \pkg{ColorBrewer.org} tool \citep{vcd:Harrower+Brewer:2003},
an online tool for selecting color palettes for maps, providing sets of colors
for coding various types of information. Here, we take a similar approach
and describe strategies for the choice of color palettes for categorical and numerical
data. Following \cite{vcd:Brewer:1999} and \cite{vcd:Harrower+Brewer:2003}, we distinguish
three types of palettes: qualitative, sequential and diverging. The first is tailored
for coding categorical information and the latter two are aimed at numerical variables.
Unlike \pkg{ColorBrewer.org}, we do not only provide fixed sets of colors (with a limited
number of colors in each set) but describe paths through a perceptually-based color space
so that the user can decide which path exactly should be taken and how many colors should
be selected. The color space employed by us is the Hue-Chroma-Luminance (HCL) color space
\citep[see][]{vcd:Ihaka:2003} and we derive explicit formulae describing the paths through
this space that correspond to different types of palettes. Given a mapping from HCL to RGB
coordinates, our formulas are extremely easy to implement and we provide such an implementation
in the \proglang{R} language \citep{vcd:R:2006} using the powerful \proglang{R} graphics
system \citep[see][]{vcd:Murrell:2006}. A flexible implementation of all palettes described
is provided in the package \pkg{vcd} \citep{vcd:Meyer+Zeileis+Hornik:2006} using the HCL
color implementation from \pkg{colorspace} \citep{vcd:Ihaka:2004}. This gives the user
both a conceptual and computational tool box for experimenting with color palettes for
a particular display.

The remainder of the paper is organized as follows: Section~\ref{sec:spaces} provides
further information on HCL and other color spaces, contrasting in particular the properties
of HSV and HCL colors. Sections~\ref{sec:qualitative}, \ref{sec:sequential}, and \ref{sec:diverging}
introduce HCL-based qualitative, sequential, and diverging palettes, respectively.
Section~\ref{sec:illustrations} shows some brief illustrations for each type of palette
in statistical graphics before Section~\ref{sec:discussion} concludes the paper with a
discussion. Some supplementary \proglang{R} code that facilitates experimentation with
the new palettes is provided in Appendix~\ref{sec:R}.

@@

@@

For choosing color palettes it is helpful to have an idea how
human color vision evolved. It has been hypothesized that it developed in three
distinct stages: 1.~perception of \textit{light/dark} contrasts (monochrome only),
2.~\textit{yellow/blue} contrasts (usually associated with our notion of warm/cold colors),
3.~\textit{green/red} contrasts (helpful for assessing the ripeness of fruit). See 
\cite{vcd:Ihaka:2003} for more details and references.

Due to these three color axes, colors are typically described as
locations in 3-dimensional spaces. However, human perception of color
does not correspond to the physiological axes above, but rather to polar
coordinates in the color plane (yellow/blue vs.~green/red) plus a third
light/dark axis. Thus, perceptually-based color spaces try to capture the
perceptual axes
\begin{enumerate}
  \item \textbf{hue} (dominant wavelength)
  \item \textbf{chroma} (colorfulness, intensity of color as compared to gray)
  \item \textbf{luminance} (brightness, amount of gray)
\end{enumerate}

A popular implementation of such a color space, available in many graphics and
statistics software packages, are \textit{HSV} (hue, saturation, value) colors.
They are a simple transformation of \textit{RGB} (red, green, blue) colors
and are defined by a triplet $(H, S, V)$ with $H \in [0, 360]$ and $S, V \in [0, 100]$.
HSV space has the shape of a single regular cone (often inflated to a regular
cylinder). Vertical sections through this space are shown in the upper panel
of Figure~\ref{fig:hsv-hcl}, depicting hue and saturation given different value levels.
Although simple to specify and easily available in many computing environments,
HSV colors have a fundamental drawback: its three dimensions map to the three
dimensions of human color perception very poorly. The three dimensions are
confounded which is most easily seen when converting the vertical sections to 
gray scale images in Figure~\ref{fig:hsv-hcl-gray}. Clearly, the brightness of
colors is not uniform over hues and saturations (given value)---therefore,
HSV colors are often not considered to be perceptually based.

To overcome these drawbacks, various color spaces have been suggested
that properly map to the perception dimensions, the most prominent of which are
the CIELUV and CIELAB spaces developed by the \cite{vcd:CIE:2004}. \cite{vcd:Ihaka:2003}
argues that CIELUV colors are typically preferred for use with emissive technologies
such as computer screens which makes them an obvious candidate for implementation
in statistical software packages. By taking polar coordinates in the UV plane of
CIELUV, \textit{HCL} (hue, chroma, luminance) colors are obtained, defined by
a triplet $(H, C, L)$ with $H \in [0, 360]$ and $C, L \in [0, 100]$. HCL space
has the shape of a distorted double cone: the admissable chroma and luminance values
depend on the hue chosen. The lower panel of Figure~\ref{fig:hsv-hcl} shows
vertical sections through this space: each of the resulting hue/chroma planes (given
luminance) is now properly balanced towards the same gray (going from black to white
with increasing luminance) which becomes obvious when converting the colors to
a gray scale is in Figure~\ref{fig:hsv-hcl-gray}. This balancing of HCL colors
gives us the opportunity to conveniently choose color palettes which code categorical
and/or numerical information by translating it to paths along the three perceptual axes.
However, some care is required for dealing with the irregular shape of the HCL space
which will be addressed in the following sections.

All palettes described are 
available in \proglang{R} \citep{vcd:R:2006} in the package \pkg{vcd} \citep{vcd:Meyer+Zeileis+Hornik:2006}
using the HCL color implementation from \pkg{colorspace} \citep{vcd:Ihaka:2004}.
Technical documentation to the \proglang{R} implementations along with a large
collection of examples is available via \verb/help("rainbow_hcl")/ that provides 
more comparisons between existing \proglang{R} palettes (based on HSV colors) and
the HCL color palettes.

@@

\cite{vcd:Ihaka:2003} describes a simple strategy for choosing such palettes: chroma
and luminance are kept fixed and only the hue is varied for obtaining different colors
which are consequently all balanced towards the same gray. As Figure~\ref{fig:hsv-hcl}
illustrates, the range of hues available depends on the combination of chroma and
luminance chosen. Figure~\ref{fig:pal-q-construct} depicts how three colors
are chosen, given $C = 50$ and $L = 70$.

Various strategies for choosing the hues in a certain palette are conceivable.
A simple and intuitive one is to use colors as metaphors for categories
(e.g., for political parties), another approach would be to use segments from
the color wheel corresponding to nearby or distant colors. The latter is shown
in Figure~\ref{fig:pal-q-examples} which depicts
examples for generating qualitative sets of colors $(H, 50, 70)$.
In the upper left panel colors from the full spectrum are used ($H = 30, 120, 210, 300$)
creating a `dynamic' set of colors. The upper right panel shows a `harmonic'
set with $H = 60, 120, 180, 240$. Warm colors (from the blue/green part of the spectrum: 
$H = 270, 230, 190, 150$) and cold colors (from the yellow/red part of the spectrum:
$H = 90, 50, 10, 330$) are shown in the lower left and right panel, respectively.

@@

In \pkg{vcd}, these palettes are available in the function
\begin{Soutput}
rainbow_hcl(n, c = 50, l = 70, start = 0, end = 360*(n-1)/n, ...)
\end{Soutput}
wher \code{n} controls the number of colors in the palette. The arguments
\code{c} and \code{l} give the fixed chroma and luminance level, respectively,
and \code{start} and \code{end} specify the range of hue angles. The function
is named after the base \proglang{R} function \code{rainbow()} which has a similar
interface but chooses colors in HSV coordinates. It computes the HCL coordinates
and transforms them to RGB coordinates by calling the \code{hcl()} function 
provided by the \pkg{colorspace} package. The \proglang{R} code for reproducing
Figure~\ref{fig:pal-q-examples} is given in the appendix.

@@

The simplest solution to this task is to employ light/dark contrasts, i.e., employ the
most basic perceptual axis. The interestingness is thus coded
by an increasing amount of gray (i.e., decreasing luminance) 
  \[ (H, 0, 90 - i \cdot 60), \]
where the hue $H$ used does not matter, chroma is set to 0 (i.e., no color), and 
luminance ranges in $[30, 90]$ avoiding the extreme colors white ($L = 100$) and 
black ($L = 0$). Instead of going linearly from light to dark gray, luminance
could also be increased nonlinearly, e.g., by some function $f(i)$ that
controls whether intensity/luminance is increased quickly or not. We found $f(i) = i^p$ 
to be a convenient transformation where the power $p$ can be varied to achieve
different degrees of non-linearity.

@@

\setkeys{Gin}{width=\textwidth}
\begin{figure}[p]
\begin{center}
<<pal-s-construct>>=
if(generate_png) {
  png(file = "pal-s-construct.png", height = 420, width = 840)
  par(mfrow = c(1, 2))
  hue.slice(260, grid.n = 301)
  lines(c(0, 100), c(90, 50), lty = 2)
  lines(c(0, 80), c(90, 30), lty = 2)
  lines(c(2.5, 2.5), c(90, 30), lty = 2)
  rect(0, 87.5, 5, 92.5, col = hcl(260, 0, 90))
  rect(47.5, 67.5, 52.5, 72.5, col = hcl(260, 50, 70))
  rect(95, 47.5, 100, 52.5, col = hcl(260, 100, 50))
  rect(37.5, 57.5, 42.5, 62.5, col = hcl(260, 40, 60))
  rect(77.5, 27.5, 82.5, 32.5, col = hcl(260, 80, 30))
  rect(0, 57.5, 5, 62.5, col = hcl(260, 0, 60))
  rect(0, 27.5, 5, 32.5, col = hcl(260, 0, 30))
  hue.slice(120, grid.n = 301)
  lines(c(0, 100), c(90, 75), lty = 2)
  lines(c(2.5, 2.5), c(90, 30), lty = 2)
  rect(0, 87.5, 5, 92.5, col = hcl(260, 0, 90))
  rect(47.5, 80, 52.5, 85, col = hcl(120, 50, 82.5))
  rect(95, 72.5, 100, 77.5, col = hcl(120, 100, 75))
  rect(0, 57.5, 5, 62.5, col = hcl(260, 0, 60))
  rect(0, 27.5, 5, 32.5, col = hcl(260, 0, 30))
  dev.off()
}
@
\includegraphics{pal-s-construct}
\caption{\label{fig:pal-s-construct} Constructing sequential palettes. The chroma/luminance
plane is shown for two hues $H = 260$ (left) and $H = 120$ (right). Colors are chosen by varying
either only luminance or both luminance and chroma.}
\end{center}
\end{figure}

\setkeys{Gin}{width=.8\textwidth}
\begin{figure}[p]
\begin{center}
<<pal-s-examples, fig=TRUE, height=5, width=6>>=
par(mfrow = c(5, 1), mar = c(0, 0, 0, 0))
pal(sequential_hcl(12, c = 0, power = 2.2))
pal(sequential_hcl(12, power = 2.2))
pal(terrain_hcl(12, c = c(65, 0), l = c(45, 95), power = c(1/3, 1.5)))
pal(heat_hcl(12, c = c(80, 30), l = c(30, 90), power = c(1/5, 1.5)))
pal(rev(heat_hcl(12, h = c(0, -100), l = c(75, 40), c = c(40, 80), power = 1)))
@
\caption{\label{fig:pal-s-examples} Examples for sequential palettes, varying
only luminance (first panel), chroma and luminance (second panel), and hue,
chroma and luminance (remaining panels).}
\end{center}
\end{figure}

In \pkg{vcd}, this strategy is implemented in the function
\begin{Soutput}
sequential_hcl(n, h = 260, c = c(80, 0), l = c(30, 90), power = 1.5, ...) 
\end{Soutput}
where the first element of \code{c} and \code{l} give the starting chroma and luminance
coordinate (by default colorful and dark) and the second element the ending coordinate
(by default light gray). The \code{power} argument implements $p$ and defaults to $1.5$.

To increase the contrast between the colors in the palette even further, the ideas from
the previous sequential palettes can also be combined with qualitative palettes by
simultaneously varying the hue as well:
  \[ (H_2 - i \cdot (H_1 - H_2), C_{\max} - i^{p_1} \cdot (C_{\max} - C_{\min}),
      L_{\max} - i^{p_2} \cdot (L_{\max} - L_{\min})). \]
One application would be a HCL-based version of heat colors that increase from a light yellow
(e.g., $(90, 30, 90)$) to a full red (e.g., $(0, 100, 50)$). To make the change in hue
visible, the chroma needs to increase rather quickly for low values of $i$ and then only
slowly for higher values of $i$. This can be achieved by choosing a power $p_1 < 1$.

In \proglang{R}, these are available in the function
\begin{Soutput}
heat_hcl(n, h = c(0, 90), c = c(100, 30), l = c(50, 90), power = c(1/5, 1), ...)
\end{Soutput}
with which the lower three rows in Figure~\ref{fig:pal-s-examples} are produced. The
\proglang{R} code reproducing this figure and stating the exact parameters used is again
provided in the appendix.

\clearpage

@@

\setkeys{Gin}{width=.8\textwidth}
\begin{figure}[p]
\begin{center}
<<pal-d-construct>>=
if(generate_png) {
  png(file = "pal-d-construct.png", height = 480, width = 534)
  hue260 <- hue.slice(260, grid.n = 301, plot = FALSE)
  hue360 <- hue.slice(360, grid.n = 301, plot = FALSE)
  mychroma <- as.numeric(colnames(hue260))
  luminance <- as.numeric(rownames(hue260))
  nc <- ncol(hue260)
  nl <- nrow(hue260)
  plot(0.5, 0.5, xlim = c(-100, 100), ylim = c(0, 100), type = "n", axes = FALSE,
       xlab = "", ylab = "", yaxs = "i", xaxs = "i", main = "")
  for(i in 1:(nc-1)) {
    rect(mychroma[i], luminance[-nl], mychroma[i] + 100/(nc-1), luminance[-1], border = hue260[,i+1], col = hue260[,i+1])
  }
  for(i in 1:(nc-1)) {
    rect(-mychroma[i], luminance[-nl], -(mychroma[i] + 100/(nc-1)), luminance[-1], border = hue360[,i+1], col = hue360[,i+1])
  }
  axis(1, labels = FALSE, at = c(-100, 0, 100))
  axis(2, at = c(0, 30, 50, 60, 70, 90, 100), labels = c(0, 30, 50, 60, 70, 90, 100))
  axis(3, at = -4:4*25, labels=c(4:0*25, 1:4*25))
  axis(4, at = c(0, 30, 50, 60, 70, 90, 100), labels = c(0, 30, 50, 60, 70, 90, 100))
  mtext(c("hue = 0", "hue = 260"), side = 3, at = c(-50, 50), line = 3, cex = 1.2)
  mtext("chroma", side = 3, at = 0, line = 2)
  mtext("luminance", side = 2, at = 50, line = 2)
  mtext("luminance", side = 4, at = 50, line = 2)
  lines(c(-100, 0, 100), c(50, 90, 50), lty = 2)
  rect(-100, 47.5, -90, 52.5, col = hcl(0, 100, 50))
  rect(-55, 67.5, -45, 72.5, col = hcl(0, 50, 70))
  rect(-5, 87.5, 5, 92.5, col = hcl(260, 0, 90))
  rect(45, 67.5, 55, 72.5, col = hcl(260, 50, 70))
  rect(90, 47.5, 100, 52.5, col = hcl(260, 100, 50))
  lines(c(-80, 0, 80), c(30, 90, 30), lty = 2)
  rect(-85, 27.5, -75, 32.5, col = hcl(0, 80, 30))
  rect(-45, 57.5, -35, 62.5, col = hcl(0, 40, 60))
  rect(-5, 87.5, 5, 92.5, col = hcl(260, 0, 90))
  rect(35, 57.5, 45, 62.5, col = hcl(260, 40, 60))
  rect(75, 27.5, 85, 32.5, col = hcl(260, 80, 30))
  dev.off()
}
@
\includegraphics{pal-d-construct}
\caption{\label{fig:pal-d-construct} Constructing diverging palettes. The chroma/luminance
plane is shown back to back hues $H = 0$ and $H = 260$. Colors are chosen by simultaneously 
varying luminance and chroma.}
\end{center}
\end{figure}

\setkeys{Gin}{width=.8\textwidth}
\begin{figure}[p]
\begin{center}
<<pal-d-examples, fig=TRUE, height=4, width=6>>=
par(mfrow = c(4, 1), mar = c(0, 0, 0, 0))
pal(diverge_hcl(7))
pal(diverge_hcl(7, c = 100, l = c(50, 90), power = 1))
pal(diverge_hcl(7, h = c(130, 43), c = 100, l = c(70, 90)))
pal(diverge_hcl(7, h = c(180, 330), c = 59, l = c(75, 95)))
@
\caption{\label{fig:pal-d-examples} Examples for diverging palettes with different
pairs of hues and decreasing luminance contrasts.}
\end{center}
\end{figure}

Given sequential palettes, deriving diverging palettes is easy: two different hues are
chosen for adding color to the same amount of `gray' at a given intensity $|i|$.
Figure~\ref{fig:pal-d-construct} shows the chroma/luminance plane back to back for the
hues $H = 0$ and 260 with two different paths---giving slightly different emphasis on
luminance or chroma contrasts---from a full red over a neutral grey to a full blue.
As Figure~\ref{fig:pal-s-construct} illustrates, the pair of hues should be chosen carefully
because the admittable values of in the chrome/luminance plane differ across hues.
Clearly, for deriving symmetric palettes, only colors from the intersection of the
admittable chroma/luminance planes can be used. The particular hues $H = 0$ and $260$
used in Figure~\ref{fig:pal-s-construct} were chosen because they have rather similar
chroma/luminance planes, allowing for both large chroma and luminance contrasts.

In \pkg{vcd}, the following function is provided:
\begin{Soutput}
diverge_hcl(n, h = c(260, 0), c = 80, l = c(30, 90), power = 1.5, ...) 
\end{Soutput}
which has the same arguments as \code{sequential_hcl()} but takes a pair of hues \code{h}.

Figure~\ref{fig:pal-d-examples} shows various examples of conceivable combinations of
hue, chroma and luminance. The first palette uses a broader range on the luminance axis
whereas the others mostly rely on chroma contrasts.

\section{Illustrations} \label{sec:illustrations}
